---
title: "02-R_Analysis"
output: html_document
date: "2025-10-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r, message=FALSE, warning=FALSE, results='hide'}
library(fs)
library(readr)
library(ggplot2)
library(knitr)
library(dplyr)
```

## Reading in Our Data

```{r}
data <- read_csv(path_wd("..", "data", "processed", "odi_bbb_recent.csv"))
```




## Time/Event Series Structure

The ball-by-ball (bbb) cricket data is inherently an event sequence structure which we can index by balls or overs. In this section I aim to demonstrate that structure and investigate any interesting features.

```{r}
data %>%
  mutate(over = ceiling(ball)) %>%
  group_by(match_id, innings, over) %>%
  summarise(over_runs = sum(runs_off_bat + extras, na.rm = TRUE), .groups = "drop") %>%
  group_by(innings, over) %>%
  summarise(avg_runs = mean(over_runs, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = over, y = avg_runs, color = factor(innings))) +
  geom_line(size = 1) +
  geom_vline(xintercept = 10, linetype = "dotted", color = "red", size = 1) +
  geom_vline(xintercept = 40, linetype = "dotted", color = "red", size = 1) +
  labs(
    title = "Average Runs per Over by Innings",
    x = "Over", 
    y = "Average Runs", 
    color = "Innings"
  ) +
  theme_minimal()



```

Here we can clearly see the event series structure. There are 3 sections in an ODI cricket game, referring to the restrictions placed on the fielding team. In the first 'Powerplay' the batting team have an advantage, making it easier for them to get boundaries (4s and 6s). We can see this by the slightly increased average runs per over for the first 10 overs (once they get their eye in). We can also see that the average runs per over generally increases with time. At 40 overs the second 'Powerplay' occurs which gives the bowling team the advantage. This period is sometimes referred to as the death and despite the fact that the bowling team has an advantage, the run rate steeply increases, particularly in the first innings. This is because they are taking more risks to attempt to set a high a target for the other team as possible. It also increases in the second innings but less so as they can afford to be more conservative as they know what goal they need to reach.

### Risk taking throughout the match

One way we can attempt to quantify risk taking/ shifts in strategy is by looking at the distribution of boundaries throughout the game. This demonstrates the underlying structure of the previous plot, a higher proportion of boundaries in the first and last 10 overs.

```{r}

data %>%
  mutate(over = ceiling(ball)) %>%
  group_by(over) %>%
  summarise(
    total_balls = n(),
    boundaries = sum(runs_off_bat %in% c(4, 6), na.rm = TRUE)
  ) %>%
  mutate(boundary_rate = boundaries / total_balls) %>%
  ggplot(aes(x = over, y = boundary_rate)) +
  geom_vline(xintercept = 10, linetype = "dashed", color = "red", size = 1) +
  geom_vline(xintercept = 40, linetype = "dashed", color = "red", size = 1) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Proportion of Balls that are Boundaries by Over",
       x = "Over",
       y = "Boundary Proportion") +
  theme_minimal()



```
Another way to see that this is 'risky' is by looking at the distribution of dismissals. Here we can see that the you are slightly more likely to be dismissed in the first 10 overs, due to the risk taking, however this is offset by having your best batters bat first. In the last 10 overs it increases massively as we have worse batters taking more risks.

```{r}
wickets_data <- data %>%
  filter(!is.na(player_dismissed)) %>%
  mutate(over = ceiling(ball))

ggplot(wickets_data, aes(x = over)) +
  geom_histogram(aes(y = after_stat(density)),
                 binwidth = 1,
                 fill = "red", color = "black", alpha = 0.6) +
  labs(title = "Density of Wickets by Over",
       x = "Over", y = "Density of Wickets") +
  geom_vline(xintercept = c(10, 40), linetype = "dotted", color = "blue") +
  theme_minimal()
```

Lets combine what we have learnt so far and see if we can see it to an individual game.


One way to visualise the event series popularised by ESPNcricinfo [2] is the 'worm chart'.

```{r}
data <- data %>%
  mutate(over = ceiling(ball))

data <- data %>%
  mutate(total_runs = runs_off_bat + ifelse(is.na(extras), 0, extras))

worm_data <- data %>%
  group_by(match_id, innings, over) %>%
  summarise(runs = sum(total_runs, na.rm = TRUE), .groups = "drop") %>%
  group_by(match_id, innings) %>%
  mutate(cum_runs = cumsum(runs))



ggplot(filter(worm_data, match_id == 749781),
       aes(x = over, y = cum_runs, colour = factor(innings))) +
  geom_line(size = 1) +
  labs(title = "Worm Chart - New Zealand vs Sri Lanka",
       x = "Over", y = "Cumulative Runs",
       colour = "Innings")


```
Here we can see that the team batting first's (Sri Lanka) run rate (gradient) stays fairly consistent until the last 10 overs where they try to run up their score and the line steepens. For New Zealand, they take advantage of the Powerplay in the first 10 overs and then can play conservatively for the next 30 Overs winning with 6 Overs to spare. Lets see if this is reflected in the distribution of wickets. We would expect Sri Lanka to pick up a lot of wickets in the last 10 overs as they play riskier.


```{r}
wickets_data <- data %>%
  filter(!is.na(player_dismissed)) %>%   # keep only rows where a wicket fell
  group_by(match_id, innings, over) %>%
  summarise(wickets = n(), .groups = "drop") %>%   # how many wickets in that over
  left_join(worm_data, by = c("match_id", "innings", "over"))

ggplot(filter(worm_data, match_id == 749781),
       aes(x = over, y = cum_runs, colour = factor(innings))) +
  geom_line(size = 1) +
  geom_point(data = filter(wickets_data, match_id == 749781),
             aes(x = over, y = cum_runs, colour = factor(innings)),
             shape = 1, size = 3, stroke = 1.2) +
  labs(title = "Worm Chart - New Zealand vs Sri Lanka",
       x = "Over", y = "Cumulative Runs",
       colour = "Innings") +
  theme_minimal()

```

We do indeed see increased frequency of wickets in the last 10 overs for Sri Lanka. We can also see that any plateaus of our lines (ie no runs scored), usually coincides with a wicket (a player being dismissed). This might suggest a bowler being on a hot streak.

## Powerplay Analysis

Let's take a look at the powerplay - in particular how many boundaries are scored
during the batting powerplay and how many wickets are lost. The big idea here 
being that if we were building a model for predicting match outcome as the game 
is being played, the powerplay would be a key indicator of how well a team's innings
is going.

```{r, Powerplay boundaries}
powerplay_boundaries <- data %>%
  filter(ball < 10, runs_off_bat %in% c(4, 6)) %>%  # only Powerplay & boundaries
  group_by(match_id, batting_team) %>%
  summarise(boundaries = n(), .groups = "drop")

powerplay_outcome <- data %>%
  distinct(match_id, batting_team, winner) %>%
  mutate(result = ifelse(batting_team == winner, "Won", "Lost"))

powerplay_summary <- powerplay_boundaries %>%
  left_join(powerplay_outcome, by = c("match_id", "batting_team")) %>%
  filter(!is.na(result))

```


```{r}
powerplay_summary %>%
  ggplot(aes(x = boundaries, fill = result)) +
  geom_histogram(position = "identity", alpha = 0.6, bins = 15) +
  scale_fill_manual(values = c("Won" = "green", "Lost" = "red")) +
  labs(title = "Boundaries in Powerplay vs Match Outcome",
       x = "Number of Boundaries in First 10 Overs",
       y = "Number of Innings",
       fill = "Match Result") +
  theme_minimal()

```
This distribution shows very clearly that teams that score more boundaries in the
*batting powerplay* tend to win more matches - the green distribution shows teams that
win matches. Therefore boundaries during the powerplay is likely to be a useful
indicator of which team will win.



## References

[1] A. Kaluarachchi and S. V. Aparna, "CricAI: A classification based tool to predict the outcome in ODI cricket," 2010 Fifth International Conference on Information and Automation for Sustainability, Colombo, Sri Lanka, 2010, pp. 250-25

[2] [ESPNcricinfo](https://www.espncricinfo.com/series/west-indies-in-england-2025-1448331/england-vs-west-indies-1st-odi-1448343/match-statistics)

[3] [dplyr cheat sheet](https://posit.co/wp-content/uploads/2022/10/data-transformation.pdf)

[4] [R for Data Science (Wickham & Grolemund)](https://r4ds.had.co.nz/transform.html)
