---
title: "02-R_Analysis"
output: html_document
date: "2025-10-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r, message=FALSE, warning=FALSE, results='hide'}
library(fs)
library(readr)
library(ggplot2)
library(knitr)
library(dplyr)
```

## Reading in Our Data

```{r}
data <- read_csv(path_wd("..", "data", "processed", "odi_bbb_recent.csv"))
```




## Time/Event Series Structure

The ball-by-ball (bbb) cricket data is inherently an event sequence structure which we can index by balls or overs. In this section I aim to demonstrate that structure and investigate any interesting features.

```{r}
data %>%
  mutate(over = ceiling(ball)) %>%
  group_by(match_id, innings, over) %>%
  summarise(over_runs = sum(runs_off_bat + extras, na.rm = TRUE), .groups = "drop") %>%
  group_by(innings, over) %>%
  summarise(avg_runs = mean(over_runs, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = over, y = avg_runs, color = factor(innings))) +
  geom_line(size = 1) +
  geom_vline(xintercept = 10, linetype = "dotted", color = "red", size = 1) +
  geom_vline(xintercept = 40, linetype = "dotted", color = "red", size = 1) +
  labs(
    title = "Average Runs per Over by Innings",
    x = "Over", 
    y = "Average Runs", 
    color = "Innings"
  ) +
  theme_minimal()



```

Here we can clearly see the event series structure. There are 3 sections in an ODI cricket game, referring to the restrictions placed on the fielding team. In the first 'Powerplay' the batting team have an advantage, making it easier for them to get boundaries (4s and 6s). We can see this by the slightly increased average runs per over for the first 10 overs (once they get their eye in). We can also see that the average runs per over generally increases with time. At 40 overs the second 'Powerplay' occurs which gives the bowling team the advantage. This period is sometimes referred to as the death and despite the fact that the bowling team has an advantage, the run rate steeply increases, particularly in the first innings. This is because they are taking more risks to attempt to set a high a target for the other team as possible. It also increases in the second innings but less so as they can afford to be more conservative as they know what goal they need to reach.

### Risk taking throughout the match

One way we can attempt to quantify risk taking/ shifts in strategy is by looking at the distribution of boundaries throughout the game. This demonstrates the underlying structure of the previous plot, a higher proportion of boundaries in the first and last 10 overs.

```{r}

data %>%
  mutate(over = ceiling(ball)) %>%
  group_by(over) %>%
  summarise(
    total_balls = n(),
    boundaries = sum(runs_off_bat %in% c(4, 6), na.rm = TRUE)
  ) %>%
  mutate(boundary_rate = boundaries / total_balls) %>%
  ggplot(aes(x = over, y = boundary_rate)) +
  geom_vline(xintercept = 10, linetype = "dashed", color = "red", size = 1) +
  geom_vline(xintercept = 40, linetype = "dashed", color = "red", size = 1) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Proportion of Balls that are Boundaries by Over",
       x = "Over",
       y = "Boundary Proportion") +
  theme_minimal()



```
Another way to see that this is 'risky' is by looking at the distribution of dismissals. Here we can see that the you are slightly more likely to be dismissed in the first 10 overs, due to the risk taking, however this is offset by having your best batters bat first. In the last 10 overs it increases massively as we have worse batters taking more risks.

```{r}
wickets_data <- data %>%
  filter(!is.na(player_dismissed)) %>%
  mutate(over = ceiling(ball))

ggplot(wickets_data, aes(x = over)) +
  geom_histogram(aes(y = after_stat(density)),
                 binwidth = 1,
                 fill = "red", color = "black", alpha = 0.6) +
  labs(title = "Density of Wickets by Over",
       x = "Over", y = "Density of Wickets") +
  geom_vline(xintercept = c(10, 40), linetype = "dotted", color = "blue") +
  theme_minimal()
```

Lets combine what we have learnt so far and see if we can see it to an individual game.


One way to visualise the event series popularised by ESPNcricinfo [2] is the 'worm chart'.

```{r}
data <- data %>%
  mutate(over = ceiling(ball))

data <- data %>%
  mutate(total_runs = runs_off_bat + ifelse(is.na(extras), 0, extras))

worm_data <- data %>%
  group_by(match_id, innings, over) %>%
  summarise(runs = sum(total_runs, na.rm = TRUE), .groups = "drop") %>%
  group_by(match_id, innings) %>%
  mutate(cum_runs = cumsum(runs))



ggplot(filter(worm_data, match_id == 749781),
       aes(x = over, y = cum_runs, colour = factor(innings))) +
  geom_line(size = 1) +
  labs(title = "Worm Chart - New Zealand vs Sri Lanka",
       x = "Over", y = "Cumulative Runs",
       colour = "Innings")


```
Here we can see that the team batting first's (Sri Lanka) run rate (gradient) stays fairly consistent until the last 10 overs where they try to run up their score and the line steepens. For New Zealand, they take advantage of the Powerplay in the first 10 overs and then can play conservatively for the next 30 Overs winning with 6 Overs to spare. Lets see if this is reflected in the distribution of wickets. We would expect Sri Lanka to pick up a lot of wickets in the last 10 overs as they play riskier.


```{r}
wickets_data <- data %>%
  filter(!is.na(player_dismissed)) %>%   # keep only rows where a wicket fell
  group_by(match_id, innings, over) %>%
  summarise(wickets = n(), .groups = "drop") %>%   # how many wickets in that over
  left_join(worm_data, by = c("match_id", "innings", "over"))

ggplot(filter(worm_data, match_id == 749781),
       aes(x = over, y = cum_runs, colour = factor(innings))) +
  geom_line(size = 1) +
  geom_point(data = filter(wickets_data, match_id == 749781),
             aes(x = over, y = cum_runs, colour = factor(innings)),
             shape = 1, size = 3, stroke = 1.2) +
  labs(title = "Worm Chart - New Zealand vs Sri Lanka",
       x = "Over", y = "Cumulative Runs",
       colour = "Innings") +
  theme_minimal()

```

We do indeed see increased frequency of wickets in the last 10 overs for Sri Lanka. We can also see that any plateaus of our lines (ie no runs scored), usually coincides with a wicket (a player being dismissed). This might suggest a bowler being on a hot streak.

## Team Performance Analysis

This analysis is to get a basic understanding of which teams generally perform than others. As these metrics are on different scales (runs have mean ~220 when wickets have an average of ~7) we must standardise these values so not all metrics are the same colour in their respective column. The way we will standardise is by subtracting the mean and dividing by the standard deviation. We must also reverse the colours for the wickets lost column, as in this case the less wickets lost the better.

Here I have chosen to do green is 'better' and red is 'worse', so if say a team scores more runs than the average, their cell will be a darker green. The idea here is to have a look at which teams are better and where - a team that has a green row is typically good, and a team that has a red row is pretty bad.

```{r}
team_performance <- data %>% #make data set
  group_by(match_id, innings, batting_team) %>%
  summarise(runs = sum(runs_off_bat),wickets = sum(!is.na(player_dismissed)),boundaries = sum(runs_off_bat %in% c(4, 6)),.groups = "drop") %>%
  group_by(batting_team) %>%
  summarise(avgruns = mean(runs),avgwickets = mean(wickets),avgboundaries = mean(boundaries),n_innings = n(),.groups = "drop")

team_heatmap_data <- team_performance %>%
  select(batting_team, avgruns, avgwickets, avgboundaries) %>%
  pivot_longer(cols = c(avgruns, avgwickets, avgboundaries),names_to = "metric",values_to = "value")
```

```{r}
team_heatmap_data <- team_heatmap_data %>%
  group_by(metric) %>%
  mutate(standardized_value = if_else(metric == "avgwickets", -(value - mean(value))/sd(value),(value - mean(value)) / sd(value))) %>% #standardise metric by subtracting mean and dividing by standard deviation
  ungroup() #reverse for wickets as less wickets is better

ggplot(team_heatmap_data, aes(x = metric, y = batting_team, fill = standardized_value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.1f", value)), color = "black", size = 3) +
  scale_fill_gradient2(low = "red", mid = "white", high = "darkgreen", midpoint = 0) +
  labs(title = "Average Team Performance Across Key Metrics",subtitle = "Green = better performance, Red = worse performance (relative to average)",x = "Performance Metric",y = "Team",fill = "Standardized\nValue") +
  scale_x_discrete(labels = c("avgboundaries" = "Boundaries","avgruns" = "Runs","avgwickets" = "Wickets Lost"))
```

This plot is to have a look at which teams perform better than others. Can see the strongest teams are England, India and Australia, closely followed by South Africa. The worst team in terms of the metrics on the chart is Papua New Guinea, who have the worst performance in each metric.

# Match Momentum Analysis 


This plot is to have a look at which teams perform better than others. Can see the strongest teams are England, India and Australia, closely followed by South Africa. The worst team in terms of the metrics on the chart is Papua New Guinea, who have the worst performance in each metric.

## Scoring Streaks and Momentum

We now look into momentum. If a team scores well in one over, this is asking the question will that carry on into the next.

After looking online, I see generally 7-9 runs in an over is a good over. We can confirm this as if a batting team averages 8 runs in an over in 50 overs they will end up with 400 runs, being a very good score. As we are looking at momentum gained after a good over, we will classify a big over as an over with 10+ runs.

```{r}
over_runs <- data %>% #get runs per over
  group_by(match_id, innings, over) %>%
  summarise(runs_in_over = sum(runs_off_bat),.groups = "drop") %>%
  arrange(match_id, innings, over) %>%
  group_by(match_id, innings) %>%
  mutate(next_over_runs = lead(runs_in_over),prev_over_runs = lag(runs_in_over)) %>%
  ungroup()

scoring_comparison <- over_runs %>% #compare what happens after big overs vs normal overs
  filter(!is.na(next_over_runs)) %>%
  mutate(over_type = case_when(runs_in_over >= 10 ~ "Big Over (10+)",runs_in_over <= 3 ~ "Quiet Over (0-3)",TRUE ~ "Normal Over (4-7)"))
```

```{r}
scoring_comparison %>%
  ggplot(aes(x = over_type, y = next_over_runs, fill = over_type)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Big Over (10+)" = "yellow","Normal Over (4-9)" = "white","Quiet Over (0-3)" = "blue")) +
  labs(title = "Runs in Next Over Based on Current Over Performance",x = "Current Over Type",y = "Runs in Next Over",fill = "Over Type")

```
We can see there is a difference for runs in an over depending on the past overs run count, which could signal that momentum has an effect, but could also be explained by a mismatched game (team being much better than opponent, meaning on average their runs per over would be higher).

## Winning Streaks Across Matches

A natural progression from single match momentum is to analyse momentum over multiple matches.

For analysis, we are going to prepare a data set arranged in teams, that will have a binary classification win = 1, lose = 0 and then make a score for the outcome of the previous 3 or so games. The charts we will make are win percentage given outcome of last match, and winning percentage given outcomes of last 3 matches.

```{r}
match_results <- data %>% #getting match results for each team in chronological order
  distinct(match_id, start_date, team1, team2, winner) %>%
  filter(!is.na(winner)) %>% #remove any non-conclusive matches
  arrange(start_date)

team_match_records <- bind_rows(match_results %>% 
    select(match_id, start_date, team = team1, winner),match_results %>% 
    select(match_id, start_date, team = team2, winner)) %>%
  arrange(team, start_date) %>% #so that each row represents each record
  mutate(won = ifelse(team == winner, 1, 0)) #binary win = 1, lose = 0

team_streaks <- team_match_records %>% #calculate score for previous matches
  group_by(team) %>%
  mutate(prev_result = lag(won),prev_2_results= lag(won) + lag(won, 2),prev_3_results= lag(won) + lag(won, 2) + lag(won, 3)) %>%
  filter(!is.na(prev_result)) %>%
  ungroup()

glimpse(team_streaks)
```
As an introduction to multi match momentum, we will first start with the probability a team wins a game, given the outcome of their previous game. We do this by setting the 

```{r}
team_streaks %>%
  mutate(previous_match = ifelse(prev_result == 1, "Won Last Match", "Lost Last Match")) %>%
  group_by(previous_match) %>%
  summarise(win_rate = mean(won),n = n(),.groups = "drop") %>%
  ggplot(aes(x = previous_match, y = win_rate, fill = previous_match)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.1f%%", win_rate * 100)), size = 5) +
  scale_fill_manual(values = c("Won Last Match" = "green", "Lost Last Match" = "red")) +
  labs(x = "Previous Match Result",y = "Win Rate in Current Match")
```
This plot doesn't seem overly useful to us, as it doesn't really tell us anything. Looking at the binary 'win match/lose match', every game is included in this calculation as every game is either won or lost. One possible interpretation could be that it shows the difference in team quality, how much better the good teams are compared to the bad ones. Still, it is a simple easy to understand plot so I thought I may as well include it.

We can try and improve the plot by extending it to winning probability given a winning streak. This could be 

```{r}
team_streaks %>% #
  filter(!is.na(prev_3_results)) %>%
  mutate(streak = case_when(prev_3_results == 3 ~ "Won last 3",prev_3_results == 0 ~ "Lost last 3",TRUE ~ "Mixed recent form")) %>% #3 wins = 1+1+1=3, 3 losses = 0+0+0=0
  group_by(streak) %>%
  summarise(win_rate = mean(won),n = n(),.groups = "drop") %>%
  ggplot(aes(x = reorder(streak, win_rate), y = win_rate, fill = streak)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.1f%%", win_rate * 100)), vjust = -0.5, size = 4) +
  coord_flip() +
  scale_fill_manual(values = c("Won last 3" = "green", "Mixed recent form" = "yellow","Lost last 3" = "red")) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.7)) +
  labs(title = "Win Rate Based on Recent Form (Last 3 Matches)",x = "Recent Form",y = "Win Rate in Current Match")
```
This can be viewed as winning streaks having an effect on winning probability, and hence supporting the idea that momentum is significant, though again it could also be interpreted as if a team is on a winning streak, they may just be a good team who wins more on average. This plot tells us more than the previous, though it definitely isn't conclusive.

## Powerplay Analysis

Let's take a look at the powerplay - in particular how many boundaries are scored
during the batting powerplay and how many wickets are lost. The big idea here 
being that if we were building a model for predicting match outcome as the game 
is being played, the powerplay would be a key indicator of how well a team's innings
is going.

```{r, Powerplay boundaries}
powerplay_boundaries <- data %>%
  filter(ball < 10, runs_off_bat %in% c(4, 6)) %>%  # only Powerplay & boundaries
  group_by(match_id, batting_team) %>%
  summarise(boundaries = n(), .groups = "drop")

powerplay_outcome <- data %>%
  distinct(match_id, batting_team, winner) %>%
  mutate(result = ifelse(batting_team == winner, "Won", "Lost"))

powerplay_summary <- powerplay_boundaries %>%
  left_join(powerplay_outcome, by = c("match_id", "batting_team")) %>%
  filter(!is.na(result))

```


```{r}
powerplay_summary %>%
  ggplot(aes(x = boundaries, fill = result)) +
  geom_histogram(position = "identity", alpha = 0.6, bins = 15) +
  scale_fill_manual(values = c("Won" = "green", "Lost" = "red")) +
  labs(title = "Boundaries in Powerplay vs Match Outcome",
       x = "Number of Boundaries in First 10 Overs",
       y = "Number of Innings",
       fill = "Match Result") +
  theme_minimal()

```
This distribution shows very clearly that teams that score more boundaries in the
*batting powerplay* tend to win more matches - the green distribution shows teams that
win matches. Therefore boundaries during the powerplay is likely to be a useful
indicator of which team will win.



## References

[1] A. Kaluarachchi and S. V. Aparna, "CricAI: A classification based tool to predict the outcome in ODI cricket," 2010 Fifth International Conference on Information and Automation for Sustainability, Colombo, Sri Lanka, 2010, pp. 250-25

[2] [ESPNcricinfo](https://www.espncricinfo.com/series/west-indies-in-england-2025-1448331/england-vs-west-indies-1st-odi-1448343/match-statistics)

[3] [dplyr cheat sheet](https://posit.co/wp-content/uploads/2022/10/data-transformation.pdf)

[4] [R for Data Science (Wickham & Grolemund)](https://r4ds.had.co.nz/transform.html)
